---
title: "Getting Started with GRAFS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with GRAFS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What are GRAFS diagrams?

GRAFS (General Representation of Agro-Food Systems) diagrams are flow diagrams
that visualize nitrogen (or other nutrient) budgets in agricultural systems.
They show the flows between compartments such as cropland, grassland, livestock,
and human population, making it possible to understand the structure and
efficiency of agro-food systems at regional scales.

The methodology was developed by Billen et al. (2014), Lassaletta et al.
(2015), and Le Noe et al. (2017).

This package automates the creation of these diagrams by:

1. Reading flow data from a CSV file
2. Populating a draw.io XML template with proportional arrow widths
3. Exporting the result as PNG images using the draw.io CLI

## Installation

### Prerequisites

The GRAFS package requires the **draw.io desktop application** to be installed
on your system. This is used to render the XML diagrams as PNG images.

Download draw.io from:
<https://github.com/jgraph/drawio-desktop/releases>

Or install via package manager:

- **Linux**: `sudo snap install drawio`
- **macOS**: `brew install --cask drawio`, then
  `chmod +x /Applications/draw.io.app/Contents/MacOS/draw.io`
- **Windows**: Download and run the installer from the releases page

### Installing the R package

```{r install}
# Install dependencies
install.packages(c("xml2", "readr", "processx"))

# Install GRAFS from a local tarball
install.packages("path/to/GRAFS_0.2.0.tar.gz", repos = NULL, type = "source")
```

### Checking your setup

After installation, verify everything is configured correctly:

```{r check}
library(GRAFS)
check_setup()
```

This will report whether draw.io is found, all R dependencies are installed,
and the bundled data files are accessible.

## Basic usage

### Locating input files

The package ships with example data for Spanish provinces:

```{r inputs}
library(GRAFS)

# Bundled input files
csv <- system.file("extdata", "GRAFS_spain_data.csv.gz", package = "GRAFS")
xml <- system.file("templates", "grafs_auto_v18.xml", package = "GRAFS")
ids <- system.file("extdata", "GRAFS_arrows_ids.csv", package = "GRAFS")

# Find draw.io executable
drawio <- find_drawio()
```

### Generating a diagram

```{r basic}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "Albacete",
  periods = list(list(years = 1930:1931))
)
```

This creates:

- `./my_output/xml/GRAFS_Albacete_P1_MEAN_1930-1931.xml` (the populated template)
- `./my_output/png/GRAFS_Albacete_P1_MEAN_1930-1931.png` (the rendered diagram)

The diagram shows the **mean** of all years in the period. Arrow widths are
proportional to flow values.

### Processing multiple regions

Pass a vector of region names to generate diagrams for each:

```{r multi-region}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = c("Albacete", "Madrid", "Barcelona"),
  periods = list(list(years = 2000:2005))
)
```

### Processing multiple periods

Pass multiple period specifications as a list:

```{r multi-period}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "Albacete",
  periods = list(
    list(years = 1900:1910),
    list(years = 1950:1960),
    list(years = 2000:2010)
  )
)
```

## Comparing periods

To visualize how flows changed between two time periods, add a `prev_years`
element to a period specification:

```{r comparison}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "Albacete",
  periods = list(
    list(years = 2011:2015, prev_years = 1990:1994)
  )
)
```

This generates a diagram where:

- Arrow widths represent the **current period** (2011-2015) mean values
- **Change bubbles** show the percentage change compared to the baseline
  (1990-1994)
- Blue bubbles (default) indicate increases
- Green bubbles (default) indicate decreases

You can customize the colors:

```{r colors}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "Albacete",
  periods = list(
    list(years = 2011:2015, prev_years = 1990:1994)
  ),
  increase_color = "#FF6B6B",
  decrease_color = "#4ECDC4"
)
```

## Controlling arrow widths

Arrow widths scale proportionally between 1 (minimum) and `max_width_arrows`
(maximum). The `val_max_width` parameter sets the data value that corresponds
to the maximum width:

```{r widths}
# For a region with small values (e.g., a single municipality)
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "Albacete",
  periods = list(list(years = 1930:1931)),
  max_width_arrows = 15,
  val_max_width = 100,
  decimals = 1
)
```

- `max_width_arrows = 15`: thinner maximum arrows
- `val_max_width = 100`: a flow value of 100 gets maximum width
- `decimals = 1`: show one decimal place (useful for small values)

## Preparing your own data

### CSV format

The input CSV must have these columns:

| Column       | Type      | Description |
|------------- |-----------|-------------|
| `province`   | character | Region name (used to filter and label diagrams) |
| `year`       | integer   | Year of the observation |
| `label`      | character | Flow label matching the template (e.g., `{CROPS_TO_POP}`) |
| `data`       | numeric   | Flow value (dot as decimal separator) |
| `align`      | character | Text alignment: `"L"` or `"R"` |
| `arrowColor` | character | Optional hex color (e.g., `#FF0000`), or `NA` |

Example rows:

```
province,year,label,data,align,arrowColor
Albacete,2010,{CROPS_TO_POP},245.5,L,NA
Albacete,2010,{LIVESTOCK_TO_HUMAN},180.3,R,NA
Albacete,2010,{PROVINCE_NAME},Albacete,L,NA
```

### Special labels

Two labels have special handling:

- `{PROVINCE_NAME}`: Replaced with the region name text (not an arrow)
- `{YEAR}`: Replaced with a formatted year range like `mean_2011-2015`

### Arrow ID mapping

The `arrows_csv` file maps flow labels to internal draw.io element IDs. The
bundled `GRAFS_arrows_ids.csv` maps 25 standard GRAFS flows. If you modify
the draw.io template to add new arrows, you must add their IDs to this file.

### Validating your inputs

Before generating diagrams, check that your data, template, and arrow mappings
are consistent:

```{r validate}
result <- validate_inputs(csv, ids, xml, regions = "Albacete")
```

This reports:

- How many labels exist in each file
- Which data labels have arrow mappings (will render as dynamic arrows)
- Which data labels are text-only (present in template but no arrow)
- Which labels are in data but not in the template (ignored)
- Whether requested regions exist in the data

## Troubleshooting

### "draw.io executable not found"

Run `check_setup()` for platform-specific installation instructions, or
provide the path manually:

```{r manual-path}
create_GRAFS(
  ...,
  drawio_exe = "/path/to/your/drawio"
)
```

### Many warnings about "Cell with label ... not found"

This is normal when your CSV contains labels that don't have corresponding
arrows in the template (e.g., intermediate calculations, area data). Use
`validate_inputs()` to see which labels are mapped and which are not.

### Arrow widths look wrong

Adjust `val_max_width` to match your data range. If your largest flow is ~500,
set `val_max_width = 500`. If all arrows look the same thickness, your
`val_max_width` is probably too large relative to your data.

## References

- Billen, G., et al. (2014). *Regional Studies*.
- Lassaletta, L., et al. (2015). *Global Biogeochemical Cycles*.
- Le Noe, J., et al. (2017). *Science of the Total Environment*.
- Rodriguez, A., et al. (2023). *Science of the Total Environment* 892: 164467.
