---
title: "Getting started with GRAFS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with GRAFS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What are GRAFS diagrams?

GRAFS (Generalized Representation of Agro-Food Systems) diagrams are flow diagrams
that visualize nitrogen (or other nutrient) budgets in agricultural systems.
They show the flows between compartments such as cropland, grassland, livestock,
and human population, making it possible to understand the structure of agro-food systems at national or regional scales.

The methodology was developed by Billen et al. (2014), Lassaletta et al.
(2015), and Le Noe et al. (2017).

This package automates the creation of these diagrams by:

1. Reading flow data from a CSV file
2. Populating a draw.io XML template with proportional arrow widths
3. Exporting the result as PNG images using the draw.io CLI

## Installation

### Prerequisites

The GRAFS package requires the **draw.io desktop application** to be installed
on your system. This is used to render the XML diagrams as PNG images.

Download draw.io from:
<https://github.com/jgraph/drawio-desktop/releases>

Or install via package manager:

- **Linux**: `sudo snap install drawio`
- **macOS**: `brew install --cask drawio`, then
  `chmod +x /Applications/draw.io.app/Contents/MacOS/draw.io`
- **Windows**: Download and run the installer from the releases page

### Installing the R package

```{r install}
# Install dependencies
install.packages(c("xml2", "readr", "processx"))

# Install GRAFS from a local tarball
install.packages("path/to/GRAFS_0.2.0.tar.gz", repos = NULL, type = "source")
```

### Checking your setup

After installation, verify everything is configured correctly:

```{r check}
library(GRAFS)
check_setup()
```

This will report whether draw.io is found, all R dependencies are installed,
and the bundled data files are accessible.

## Basic usage

### Locating input files

The package ships with example national-level data for Spain, 1990-2015
(from Rodriguez et al. 2023 -- see References):

```{r inputs}
library(GRAFS)

# Bundled input files
csv <- system.file("extdata", "GRAFS_spain_data.csv.gz", package = "GRAFS")
xml <- system.file("templates", "grafs_auto_v18.xml", package = "GRAFS")
ids <- system.file("extdata", "GRAFS_arrows_ids.csv", package = "GRAFS")

# Find draw.io executable
drawio <- find_drawio()
```

### Generating a diagram

```{r basic}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "spain",
  periods = list(list(years = 1990:1991))
)
```

This creates:

- `./my_output/xml/GRAFS_spain_P1_MEAN_1990-1991.xml` (the populated template)
- `./my_output/png/GRAFS_spain_P1_MEAN_1990-1991.png` (the rendered diagram)

The diagram shows the **mean** of all years in the period. Arrow widths are
proportional to flow values.

### Processing multiple regions

Pass a vector of region names to generate diagrams for each. The bundled
example data only has one region (`"spain"`), but if your own CSV has
multiple regions in its `province` column, you can process them all at once:

```{r multi-region}
create_GRAFS(
  csv_inputs = "path/to/your_data.csv",
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = c("region_1", "region_2", "region_3"),
  periods = list(list(years = 2000:2005))
)
```

### Processing multiple periods

Pass multiple period specifications as a list:

```{r multi-period}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "spain",
  periods = list(
    list(years = 1990:1995),
    list(years = 2000:2005),
    list(years = 2010:2015)
  )
)
```

## Comparing periods

To visualize how flows changed between two time periods, add a `prev_years`
element to a period specification:

```{r comparison}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "spain",
  periods = list(
    list(years = 2011:2015, prev_years = 1990:1994)
  )
)
```

This generates a diagram where:

- Arrow widths represent the **current period** (2011-2015) mean values
- **Change bubbles** show the percentage change compared to the baseline
  (1990-1994)
- Blue bubbles (default) indicate increases
- Green bubbles (default) indicate decreases

If you don't need this comparison, delete the whole `prev_years = ...` part
(as in the "Processing multiple periods" example above, which has no
`prev_years`) -- no change bubbles will be drawn.

You can customize the colors:

```{r colors}
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "spain",
  periods = list(
    list(years = 2011:2015, prev_years = 1990:1994)
  ),
  increase_color = "#FF6B6B",
  decrease_color = "#4ECDC4"
)
```

## Controlling arrow widths

Arrow widths scale proportionally between 1 (minimum) and `max_width_arrows`
(maximum). The `val_max_width` parameter sets the data value that corresponds
to the maximum width:

```{r widths}
# For a region with small values (e.g., a single municipality)
create_GRAFS(
  csv_inputs = csv,
  path_outputs = "./my_output",
  xml_base = xml,
  arrows_csv = ids,
  drawio_exe = drawio,
  regions = "spain",
  periods = list(list(years = 1990:1991)),
  max_width_arrows = 15,
  val_max_width = 100,
  decimals = 1
)
```

- `max_width_arrows = 15`: thinner maximum arrows
- `val_max_width = 100`: a flow value of 100 gets maximum width
- `decimals = 1`: show one decimal place (useful for small values)

## Preparing your own data

### CSV format

The input CSV must have these columns:

| Column       | Type      | Description |
|------------- |-----------|-------------|
| `province`   | character | Region name (used to filter and label diagrams) |
| `year`       | integer   | Year of the observation |
| `label`      | character | Flow label matching the template (e.g., `{CROPS_TO_POP}`) |
| `data`       | numeric (character for `{PROVINCE_NAME}`) | Flow value (dot as decimal separator), or the region name for `{PROVINCE_NAME}` |
| `align`      | character | Text alignment: `"L"` or `"R"` |
| `arrowColor` | character | Optional hex color (e.g., `#FF0000`), or `NA` |

Example rows:

```
province,year,label,data,align,arrowColor
spain,2010,{CROPS_TO_POP},245.5,L,NA
spain,2010,{LIVESTOCK_TO_HUMAN},180.3,R,NA
spain,2010,{PROVINCE_NAME},Spain,L,NA
```

### Special labels

Two labels have special handling:

- `{PROVINCE_NAME}`: Replaced with the region name text from the CSV's `data`
  column (not an arrow)
- `{YEAR}`: Automatically replaced with a formatted year range like
  `2011-2015`, derived from the `years` you pass in -- no CSV value needed

### Arrow ID mapping

The `arrows_csv` file maps flow labels to internal draw.io element IDs. The
bundled `GRAFS_arrows_ids.csv` maps 25 standard GRAFS flows.

You can revise the draw.io template itself -- for example, removing arrows
for flows you don't want to show. If you modify the
template to add new arrows, add their IDs to `arrows_csv`; if you remove
arrows, remove the corresponding rows from `arrows_csv` too.

### Validating your inputs

Before generating diagrams, check that your data, template, and arrow mappings
are consistent:

```{r validate}
result <- validate_inputs(csv, ids, xml, regions = "spain")
```

This reports:

- How many labels exist in each file
- Which data labels have arrow mappings (will render as dynamic arrows)
- Which data labels are text-only (present in template but no arrow)
- Which labels are in data but not in the template (ignored)
- Whether requested regions exist in the data

## Label reference

The labels used in the bundled dataset and template are nitrogen (N) budget
variables from the GRAFS methodology. This reference explains what each one
represents; it's provided by the package's original authors and may not
cover every label if you extend the template with your own.

### Facts (informational values, not flows)

| Label | Meaning |
|-------|---------|
| `{PROVINCE_NAME}` | Name of the study area |
| `{YEAR}` | The year range shown in the diagram |
| `{YEARCHANGE}` | The comparison year range shown in the diagram (when a comparison period is requested) |
| `{KM2_PROVINCE}` | Area of the study region in km² (named "province" even when the study area isn't a province) |
| `{WIDTH_MAX}` | The data value that maps to the maximum arrow width in the diagram |
| `{POPULATION}` | Population, in inhabitants |
| `{POPULATIONM}` | Population, in millions of inhabitants |
| `{GRASSMha}` | Approximate grassland area, in Mha |
| `{GREHha}` / `{GREHMha}` / `{GREHN}` | Greenhouse crops area (ha / Mha) and N quantity (GgN) |
| `{PERiha}` / `{PERiMha}` / `{PERiN}` | Permanent irrigated crops area (ha / Mha) and N production (GgN) |
| `{PERrha}` / `{PERrMha}` / `{PERrN}` | Permanent rainfed crops area (ha / Mha) and N production (GgN) |
| `{NPEiha}` / `{NPEiMha}` / `{NPEiN}` | Non-permanent irrigated crops area (ha / Mha) and N production (GgN) |
| `{NPErha}` / `{NPErMha}` / `{NPErN}` | Non-permanent rainfed crops area (ha / Mha) and N production (GgN) |
| `{MONOGLU}` / `{MONOGMLU}` | Monogastric livestock units (and in millions) |
| `{RUMIANTSLU}` / `{RUMIANTSMLU}` | Ruminant livestock units (and in millions) |
| `{LVSTCKTOTN}` | Total N in livestock (`{LIVESTOCK_TO_HUMAN}` + `{LIVESTOCK_EXPORTED}` + `{AN_LS_OTH}`); the vertical label on the livestock box |
| `{CRPLNDTOTN}` | Total N production from cropland; the vertical label on the cropland box |

### Inputs (N flowing in)

| Label | Meaning |
|-------|---------|
| `{CROP_POPIMPORT}` | N from vegetable crops imported, to feed the population |
| `{CROPS_TO_LIVESTOCK}` | N from crops to livestock (feed) |
| `{CROPS_TO_POP}` | N from the system's own crops consumed by the population |
| `{CRTOLVSTCK_R}` | N from crops to ruminant livestock |
| `{FIX_DEP_CR}` | Symbiotic N fixation and atmospheric N deposition into cropland |
| `{FIX_DEP_GRASS}` | Symbiotic N fixation and atmospheric N deposition into grassland |
| `{FIXCR}` | N fixed by crops |
| `{FIXGR}` | N fixed by grasslands |
| `{GRASS_TO_LIVESTOCK}` | N from grassland to feed livestock |
| `{IMANOT}` | N imported for animals, other (mineral compounds, dietary supplements) |
| `{IMPHMANA}` | N in imported animal products consumed by humans |
| `{IMPORT_ANIMALCR_MONOG}` | N from imported crops to feed monogastric livestock |
| `{IMPORT_ANIMALCR_RUM}` | N from imported crops to feed ruminant livestock |
| `{IMPORT_ANIMALCR}` | Sum of the above two (imported crops to feed livestock) |
| `{LIVESTOCK_TO_CROPS}` | N manure from livestock applied to cropland |
| `{LIVESTOCK_TO_GRASS}` | N manure from livestock applied to grassland |
| `{LIVESTOCK_TO_HUMAN}` | N in livestock consumed by humans (needs minus imported animal products) |
| `{ORGOT}` | N from other organic sources fertilizing cropland (e.g. urban waste) |
| `{SYF_GRASS}` | N from synthetic fertilizers applied to grassland |
| `{SYNTHF_TOTAL}` | Total N from synthetic fertilizers applied to crops and grassland |
| `{SYNTHF}` | N from synthetic (mineral) fertilizers applied to crops |

### Outputs (N flowing out)

| Label | Meaning |
|-------|---------|
| `{AN_LS_OTH}` | N in animal losses and others (`{AN_LS}` + `{AN_OTH}`) |
| `{AN_LS}` | N in animal losses |
| `{AN_OTH}` | N in animal for other uses |
| `{CROP_EXPORT}` | N from crops exported outside the system |
| `{CROP_SURPLUS}` | N surplus from crops (deposition + fixation + synthetic + manure + other organic sources, minus feed, human consumption, losses, and exports) |
| `{CRP_LS_OTHUSES}` | Losses and other uses of N content of crops |
| `{CRP_LS}` | Losses of N content of crops |
| `{CRP_OTHUSES}` | N from crop losses from other uses |
| `{LIVESTOCK_EXPORTED}` | N in livestock exported |
| `{LIVGASLOSS}` | N losses from livestock through gaseous volatilization |
| `{LVSTCK_NOEDIBLE}` | N in non-edible material from livestock |
| `{UNKMANURE}` | N from manure with unknown destination -- after accounting for all known uses, some manure N remains unallocated, and is treated as an unknown loss (likely polluting the environment) |
| `{WASTEWATER}` | N lost in wastewater generated by the human population |

## Troubleshooting

### "draw.io executable not found"

Run `check_setup()` for platform-specific installation instructions, or
provide the path manually:

```{r manual-path}
create_GRAFS(
  ...,
  drawio_exe = "/path/to/your/drawio"
)
```

### Many warnings about "Cell with label ... not found"

This is normal when your CSV contains labels that don't have corresponding
arrows in the template (e.g., intermediate calculations, area data). Use
`validate_inputs()` to see which labels are mapped and which are not.

### Arrow widths look wrong

Adjust `val_max_width` to match your data range. If your largest flow is ~500,
set `val_max_width = 500`. If all arrows look the same thickness, your
`val_max_width` is probably too large relative to your data.

## References

- Billen, G., et al. (2014). *Regional Studies*.
- Lassaletta, L., et al. (2015). *Global Biogeochemical Cycles*.
- Le Noe, J., et al. (2017). *Science of the Total Environment*.
- Rodriguez, A., et al. (2023). *Science of the Total Environment* 892: 164467.
